Supplementary Materials

### 2021-09-30 analysis/paper/abstract.Rmd
revisions to first draft

@@

<!-- One sentence clearly stating the general problem being addressed by this particular study -->

<!-- One sentence summarising the main result (with the words "here we show" or their equivalent) -->

@@

<!-- Two or three sentences to provide a broader perspective, readily comprehensible to a scientist in any discipline, may be included in the first paragraph if the editor considers that the accessibility of the paper is significantly enhanced by their inclusion. Under these circumstances, the length of the paragraph can be yp to 300 words. (This example is 190 words without the final section, and 250 words with it).

>>>>

```{r include=FALSE}
w_count <- wordcountaddin::word_count()
```

@@

Dental calculus, mineralised plaque, has proven to contain a wealth of information 
on the dietary habits of past populations. The extraction and identification of
starch granules contained within dental calculus has already provided valuable
dietary insights from a wide range of regions and time periods, and seem to preserve 
well in the archaeological record.
<!-- One sentence clearly stating the general problem being addressed by this particular study -->
While the majority of studies have focused on microfossil extraction and
identification from archaeological remains, <!--
as well as methods of extraction and decontamination,--> few studies have 
addressed the biases of starch retention in dental calculus, and have been
limited to *in vivo* studies on modern humans and non-human primates.
<!--
and have primarily been 
conducted on modern humans, non-human primates, and archaeological remains; 
all of which have 
methodological challenges of dealing with living (or once living) organisms, 
and the variables that are difficult to control. <!-- rephrase -->

<!-- One sentence summarising the main result (with the words "here we show" or their equivalent) -->
Here, we present a multispecies oral biofilm model, which allows experimental
research to be conducted on *in vitro* dental calculus in a controlled laboratory
setting. The biofilms were exposed treatment solutions with known quantities of
dietary starches (wheat and potato) during the 25-day growth period. After this,
the starch granules were extracted from the mature biofilm (dissolution in EDTA),
and counted. We 
show that the granule counts extracted from the model dental calculus represented 
a low proportion of the total counts (ranging from 0.064% to 0.161%) exposed to 
the biofilms throughout the experiment. Additionally, we found that the ratios
of granule sizes from the extracted starch granules differed from the original
treatment solutions, with large granules (>20 $\mu$m) consistently being
under-represented.
Additionally, the size ratios of granules within starch species shifted between the input starch solutions and the granules extracted from the model calculus, with large granules (>20 $\mu$m) consistently being underrepresented. We also found a correlation
between the absolute granule counts and dry-weight of the biofilm (r = ), as well
as between the concentration (count per mg) of granules and dry-weight (r = ).

@@

<!-- Two or three sentences to provide a broader perspective, readily comprehensible to a scientist in any discipline, may be included in the first paragraph if the editor considers that the accessibility of the paper is significantly enhanced by their inclusion. -->
Our results reinforce previous *in vivo* studies suggesting that dental calculus 
presents a very small, and partly biased picture of the original dietary intake 
of starches, with an over-representation of plants producing granules smaller than
20 $\mu$ in size. The experimental model presented here is well-suited to address
the need for further validation of methods and biases associated with dietary
research on dental calculus.

Word count: `r w_count`

### 2021-09-30 analysis/supplementary-materials/supp-mat.Rmd
revisions to first draft

---
title: Supplementary Materials
output: 
  html_document:
    keep_md: true
    df_print: paged
---

```{r setup}
library(here)
library(tidyverse)
raw_data <- read_csv(here("analysis/data/raw_data/starch_counts.csv"))
```

## Metadata

**raw_counts**
<!-- insert data type for each column -->
Counts represent the absolute number of starches counted on a slide

sample: Sample number.
plate: Plate number that the sample came from.
row: Which row on the plate the sample came from.
s: Small starch count.
m: Medium starch count.
l: Large starch count.
total: Sum of s, m, and l.
treatment: Treatment solution to which the samples were exposed.
starch: Type of starch that was counted
weight: Weight of the biofilm sample
vol: Total volume of EDTA in which the sample was dissolved.
portion_slide: Proportion of the microscope slide that was counted.

@@

@@

$$ \text{corrected count} = \text{raw count} \times \frac{29}{3} \times 100 \mu l $$

## Functions

>>>>

---
title: Supplementary Materials
output: 
  html_document:
    keep_md: true
    df_print: paged
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
library(here)
library(tidyverse)
raw_data <- read_csv(here("analysis/data/raw_data/starch_counts.csv"))
```

## R Session info

```{r}
print(sessionInfo(), locale = FALSE)
```

## Metadata for raw data files

Counts represent the absolute number of starches counted on a slide

**starch_counts.csv**

| variable | description |
|----|----|
| sample | Sample number. |
| plate | Plate number that the sample came from. |
| row | Which row on the plate the sample came from. |
| s | Small starch count. |
| m | Medium starch count. |
| l | Large starch count. |
| total | Sum of s, m, and l. |
| treatment | Treatment solution to which the samples were exposed. |
| starch | Type of starch that was counted. |
| weight | Weight of the biofilm sample. |
| vol | Total volume of EDTA in which the sample was dissolved. |
| portion_slide | Proportion of the microscope slide that was counted. Total transects on slide divided by counted transects. |

**solution_counts**

| variable | description |
|----|----|
| solution | Type of starch in solution. |
| concentration | Concentration (%w/v) of starch in solution. | 
| vol_slide | Volume of solution added to slide. |
| vol_total | Total volume of solution in aliquot. |
| portion_slide | Proportion of slide that was counted. Total transects on slide divided by counted transects. |
| slide | Slide number. |
| starch | Starch type counted. |
| s | Small starch count. |
| m | Medium starch count. |
| l | Large starch count. |
| total | Sum of s, m, and l. |

## Amylase activity

Amylase activity in U/mL enzyme, where a U is mg maltose released from starch in six minutes at 36 &deg;C.

Tables containing the amylase activity results for both plates and photometric
readings.

```{r amylase-activity-results, message=FALSE}
# table of results reported in units amylase per mL enzyme (but let's be honest,
  # ...it doesn't really matter what the unit is. No activity is no activity)
source(here("analysis/scripts/amylase-assay.R"))
cols <- c("1", "2", "3")  
rows <- c("S1", "S2", "S3", "B1", "B2", "B3", "B4", "B5", "BT1", "BT2", "BT3")
plt1_ph1_result <- rbind(sal1_ph1, bmm1_ph1)
rownames(plt1_ph1_result) <- rows
plt1_ph2_result <- rbind(sal1_ph2, bmm1_ph2)
rownames(plt1_ph2_result) <- rows
plt2_ph1_result <- rbind(sal2_ph1, bmm2_ph1)
rownames(plt2_ph1_result) <- rows
plt2_ph2_result <- rbind(sal2_ph2, bmm2_ph2)
rownames(plt2_ph2_result) <- rows
```

```{r amylase-activity-tables}
plt1_ph1_result
plt1_ph2_result
plt2_ph1_result
plt2_ph2_result
```

@@

Slide transects were calculated by counting the number of transects on the 
cover slip under the microscope. This was done by starting in the bottom-left
corner, and counting the total number of full fields-of-view across the cover 
slip to the bottom-right corner. The total number of transects was 29 (verified
multiple times).

$$ \text{corrected count} = \text{raw count} \times \frac{\text{total} \space \text{transects}}{\text{counted} \space \text{transects}} \times 100 \mu l $$

@@

## Microscope images

```{r img-1, echo=FALSE, fig.cap="Image of starch granules extracted from a calculus sample"}

```
